---
title: "NHANES Dataset Description"
author: "Anthony Dumais"
date: "2026-09-30"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
library(dplyr)
library(rio)
library(here)

nhanes <- rio::import(here::here("data", "cleaned_NHANES.csv")) %>%
  mutate(
    wave = factor(
      wave_id,
      levels = c(2015, 2016, 2017),
      labels = c("2013–2014", "2015–2016", "2017–2018")
    )
  )
```

# Introduction

This report describes the unweighted characteristics of 29,400 participants from three cycles of the National Health and Nutrition Examination Survey (NHANES) [@nhanesdata]. The tables compare sample size, gender, ethnicity, and average systolic blood pressure across cycles.

# Sample size and gender

Table 1 shows the number of participants and the number and percentage of males in each survey cycle.

```{r sample-gender-table, echo=FALSE}
sample_summary <- nhanes %>%
  group_by(wave) %>%
  summarise(
    total_n = n(),
    male_n = sum(gender == "Male", na.rm = TRUE),
    male_pct = 100 * male_n / total_n,
    .groups = "drop"
  )

sample_summary %>%
  mutate(
    total_n = format(total_n, big.mark = ",", trim = TRUE),
    male = sprintf(
      "%s (%.1f%%)",
      format(male_n, big.mark = ",", trim = TRUE),
      male_pct
    )
  ) %>%
  select(wave, total_n, male) %>%
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    col.names = c("Survey cycle", "Total, n", "Males, n (%)"),
    caption = "Sample size and male participants by NHANES survey cycle"
  ) %>%
  kableExtra::kable_styling(latex_options = "hold_position") %>%
  kableExtra::column_spec(1, bold = TRUE) %>%
  kableExtra::footnote(
    general = "Unweighted counts; percentages use all participants in each cycle.",
    threeparttable = TRUE
  )
```

Across the three cycles, the dataset contains `r format(nrow(nhanes), big.mark = ",")` participants. The percentage of males ranges from `r sprintf("%.1f", min(sample_summary$male_pct))`% to `r sprintf("%.1f", max(sample_summary$male_pct))`%.

# Ethnicity

Table 2 presents the number and percentage of participants in each ethnicity category, within each survey cycle.

```{r ethnicity-table, echo=FALSE}
ethnicity_summary <- nhanes %>%
  count(wave, ethnicity_2, name = "n") %>%
  group_by(wave) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

ethnicity_summary %>%
  mutate(
    entry = sprintf(
      "%s (%.1f%%)",
      format(n, big.mark = ",", trim = TRUE),
      pct
    )
  ) %>%
  select(ethnicity_2, wave, entry) %>%
  tidyr::pivot_wider(names_from = wave, values_from = entry) %>%
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    col.names = c("Ethnicity", "2013–2014", "2015–2016", "2017–2018"),
    caption = "Ethnicity of NHANES participants by survey cycle"
  ) %>%
  kableExtra::kable_styling(
    latex_options = "HOLD_position",
    font_size = 9
  ) %>%
  kableExtra::add_header_above(
    c(" " = 1, "Survey cycle: n (%)" = 3)
  ) %>%
  kableExtra::column_spec(1, bold = TRUE) %>%
  kableExtra::footnote(
    general = "Unweighted counts; percentages are within each survey cycle."
  )
```
The White category accounts for between `r sprintf("%.1f", min(ethnicity_summary$pct[ethnicity_summary$ethnicity_2 == "White"]))`% and `r sprintf("%.1f", max(ethnicity_summary$pct[ethnicity_summary$ethnicity_2 == "White"]))`% of participants across the three cycles.

# Average systolic blood pressure

Each participant’s average systolic blood pressure (SBP) is calculated from their available readings. Table 3 summarizes these averages by survey cycle.

```{r sbp-table, echo=FALSE}
sbp_columns <- paste0("systolic_bp_", 1:4)

nhanes$average_sbp <- rowMeans(nhanes[sbp_columns], na.rm = TRUE)
nhanes$average_sbp[is.nan(nhanes$average_sbp)] <- NA_real_

sbp_summary <- nhanes %>%
  group_by(wave) %>%
  summarise(
    measured_n = sum(!is.na(average_sbp)),
    min_sbp = min(average_sbp, na.rm = TRUE),
    max_sbp = max(average_sbp, na.rm = TRUE),
    median_sbp = median(average_sbp, na.rm = TRUE),
    mean_sbp = mean(average_sbp, na.rm = TRUE),
    sd_sbp = sd(average_sbp, na.rm = TRUE),
    .groups = "drop"
  )

sbp_summary %>%
  mutate(
    measured_n = format(measured_n, big.mark = ",", trim = TRUE),
    across(
      c(min_sbp, max_sbp, median_sbp, mean_sbp, sd_sbp),
      ~ sprintf("%.1f", .x)
    )
  ) %>%
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    col.names = c("Survey cycle", "n", "Minimum", "Maximum",
                  "Median", "Mean", "SD"),
    caption = "Average systolic blood pressure by NHANES survey cycle"
  ) %>%
  kableExtra::kable_styling(
    latex_options = "HOLD_position",
    font_size = 9
  ) %>%
  kableExtra::add_header_above(
    c(" " = 2, "Average SBP (mm Hg)" = 5)
  ) %>%
  kableExtra::column_spec(1, bold = TRUE) %>%
  kableExtra::footnote(
    general = "n = participants with at least one SBP reading; SD = standard deviation. Statistics describe participants' average SBP.",
    threeparttable = TRUE
  )
```

Mean average SBP ranges from `r sprintf("%.1f", min(sbp_summary$mean_sbp))` to `r sprintf("%.1f", max(sbp_summary$mean_sbp))` mm Hg across the three cycles. Participants without any SBP readings are excluded from these summaries.

# References